---
title: Value Tree
---

------------------------------------------------------------------------

```{r, include = FALSE}
value_tree <- function(diagram, ...) {
  if (is.null(diagram) || nchar(trimws(diagram)) == 0) {
    stop("A diagram specification must be provided")
  }

  # Basic syntax check
  if (!grepl("^graph (TB|BT|RL|LR);", trimws(diagram))) {
    stop("Invalid graph specification: Must start with 'graph TB;', 'graph BT;', 'graph RL;', or 'graph LR;'")
  }

  # Attempt to create the DiagrammeR object
  tryCatch(
    {
      result <- DiagrammeR::mermaid(diagram, ...)
      return(result)
    },
    error = function(e) {
      stop(paste("Invalid graph specification:", e$message))
    }
  )
}
```

::: panel-tabset
## Standard Visualization
```{r, echo = FALSE}
value_tree(
  diagram =
    "graph LR;
   A(<B>Benefit-Risk Balance</B>)-->B(<B>Benefits</B>)
   B-->C(<B>Primary Efficacy</B>)
   B-->D(<B>Secondary Efficacy</B>)
   B-->E(<B>Quality of life</B>)
   C-->F(<B>% Success</B>)
   D-->G(<B>Mean change</B>)
   E-->H(<B>Mean change</B>)
   A-->I(<B>Risks</B>)
   I-->J(<B>Recurring AE</B>)
   I-->K(<B>Rare SAE</B>)
   I-->L(<B>Liver Toxicity</B>)
   J-->M(<B>Event rate</B>)
   K-->N(<B>% Event</B>)
   L-->O(<B>% Event</B>)
   style A fill:#7ABD7E
   style B fill:#7ABD7E
   style I fill:#7ABD7E
   style C fill:#FFE733
   style D fill:#FFE733
   style E fill:#FFE733
   style J fill:#FFE733
   style K fill:#FFE733
   style L fill:#C6C6C6
   style F fill: #FFAA1C
   style G fill: #FFAA1C
   style H fill: #FFAA1C
   style M fill: #FFAA1C
   style N fill: #FFAA1C
   style O fill: #C6C6C6
   "
)
```

## Sample Code

```{r, eval = FALSE}
value_tree(
  diagram =
    "graph LR;
   A(<B>Benefit-Risk Balance</B>)-->B(<B>Benefits</B>)
   B-->C(<B>Primary Efficacy</B>)
   B-->D(<B>Secondary Efficacy</B>)
   B-->E(<B>Quality of life</B>)
   C-->F(<B>% Success</B>)
   D-->G(<B>Mean change</B>)
   E-->H(<B>Mean change</B>)
   A-->I(<B>Risks</B>)
   I-->J(<B>Recurring AE</B>)
   I-->K(<B>Rare SAE</B>)
   I-->L(<B>Liver Toxicity</B>)
   J-->M(<B>Event rate</B>)
   K-->N(<B>% Event</B>)
   L-->O(<B>% Event</B>)
   style A fill:#7ABD7E
   style B fill:#7ABD7E
   style I fill:#7ABD7E
   style C fill:#FFE733
   style D fill:#FFE733
   style E fill:#FFE733
   style J fill:#FFE733
   style K fill:#FFE733
   style L fill:#C6C6C6
   style F fill: #FFAA1C
   style G fill: #FFAA1C
   style H fill: #FFAA1C
   style M fill: #FFAA1C
   style N fill: #FFAA1C
   style O fill: #C6C6C6
   "
)
```

## Description

**How to read**

-   A value tree is a diagram that includes all key benefits and risks identified by the stakeholder or decision-makers to be important to the decision, particular to the patient population considered.

-   The tree has no quantitative values, but rather displays outcomes valued by decision makers.

-   The benefits typically include primary and secondary clinical endpoints, and a Quality of Life measure. Occasionally it includes a measure of convenience and out-of-pocket expense.

    -   Benefits are statistically characterized at the right as means, mean change from baseline, proportions (ex.% Success), or rates.

-   The risks are typically organ toxicities or adverse events of special interest and can be statistically characterized as proportions or rates (ex. incidence per 100 subjects or per 10,000 subject-years).

-   The value tree can be used to facilitate team discussions about which outcomes to capture in a clinical study.

-   Value trees can start with many branches that can be pruned or combined to form a smaller tree that best assess the tradeoffs between key benefits and key risks.

-   If two outcomes are highly correlated, then only one should be used, so as to avoid double counting.

-   The same outcome can be specified two different ways, either a clinical endpoint (ex. blood pressure) or a patient-focused outcome (ex. ability to climb stairs). The decision makers would need to decide which one to keep.

**Key Conclusions:**

-   The size of this value tree was reduced by grouping frequently occurring adverse events into a single “recurring AE” outcome and grouping rare but serious adverse events into a single outcome.

-   While also a rare and serious, liver toxicity was colored gray because it only has a potential association with treatments.

-   Both primary and secondary efficacy clinical endpoints were included because they were not correlated. Quality of Life was included to capture what was important to the subjects.

:::
