---
title: Comorbidities (Grouped Bar Chart)
---

```{r, include = FALSE}
load("../data/comorbidities.rda")
library(ggplot2)
source("utils.R")
```

------------------------------------------------------------------------

```{r, include = FALSE}
#' Create a Grouped Bar Chart
#'
#' This function generates a grouped bar chart using ggplot2, where bars are grouped by a specified variable.
#'
#' @param data A data frame containing at least 3 variables: x-axis variable, y-axis variable, and a grouping variable.
#' @param xvar A string specifying the name of the variable to be plotted on the x-axis.
#' @param yvar A string specifying the name of the variable to be plotted on the y-axis.
#' @param groupvar A string specifying the name of the grouping variable.
#' @param chartcolors A vector of colors with the same length as the number of levels in the grouping variable.
#'
#' @return A ggplot object representing the grouped bar chart.
#'
#' @details
#' The function creates a bar chart where bars are grouped based on the specified grouping variable.
#' It uses position = "dodge" to place bars side by side within each group.
#' The y-axis line, minor grid lines, and y-axis ticks are removed for a cleaner appearance.
#'
#' @note
#' This function requires the ggplot2 package and assumes the existence of a custom theme function br_charts_theme().
#'
#' @import ggplot2
#' @export
#'
#' @examples
#' # Assuming 'comorbidities' is a data frame with appropriate columns
#' # and 'colfun()$fig4_colors' returns a vector of colors
#' grouped_barchart(
#'   data = comorbidities,
#'   xvar = "Comorbidities",
#'   yvar = "Prevalence",
#'   groupvar = "Severity",
#'   chartcolors = colfun()$fig4_colors
#' )
#'
grouped_barchart <- function(data, xvar, yvar, groupvar, chartcolors) {
  fig <- ggplot(data, aes(
    x = .data[[xvar]], y = .data[[yvar]], fill = .data[[groupvar]]
  )) +
    geom_bar(position = "dodge", stat = "identity") +
    scale_fill_manual(values = chartcolors) +
    guides(fill = guide_legend(title = paste0(groupvar, ":"))) +
    br_charts_theme(
      axis.line.y = element_blank(),
      panel.grid.minor.y = element_blank(),
      axis.ticks.y = element_blank()
    )

  fig
}
```

::: panel-tabset
## Standard Visualization

```{r, echo = FALSE, warning=FALSE}
grouped_barchart(
  data = comorbidities, xvar = "Comorbidities",
  yvar = "Prevalence", groupvar = "Severity",
  chartcolors = colfun()$fig4_colors
)
```
## Sample Code

```{r, eval = FALSE}
grouped_barchart(
  data = comorbidities, xvar = "Comorbidities",
  yvar = "Prevalence", groupvar = "Severity",
  chartcolors = colfun()$fig4_colors
)
```

## Description

**How to read**

-   The Y-axis displays the prevalence (%) of disease comorbidities in the patient population, ranging from 0 to 40%.

-   The X-axis displays five comorbidities (anxiety, …, sarcopenia), each categorized by 3 severity levels (mild, moderate, or severe) color coded according to the legend at the top.

**Key Conclusions:**

-   The graph displays the background rates of comorbidities, which is useful in designing patient preference studies. For example, patients might place high importance in a drug that reduces severe hypertension.

-   A display of background rates is also useful in determining whether a comorbidity is likely due to the drug or the disease.

:::
