---
title: Composite Outcome (Stacked Bar Chart)
---

```{r, include = FALSE}
source("utils.R")
load("../data/comp_outcome.rda")
library(dplyr)
library(ggplot2)
```

------------------------------------------------------------------------

```{r, include = FALSE}
#' Stacked Bar Chart
#'
#' @param data `dataframe` a data frame with a minimum of 4 variables named the following:
#' 1) usubjid: unique subject ID
#' 2) visit: visit ID
#' 3) trt: treatment group
#' 4) brcat: composite benefit-risk category
#' @param chartcolors `vector` a vector of colors, the same number of levels as
#' the brcat variable
#' @param xlabel `character` x label name, default is "Visit"
#'
#' @return a ggplot object
#' @export
#'
#' @examples
#' stacked_barchart(
#'   data = comp_outcome,
#'   chartcolors = colfun()$fig12_colors,
#'   xlabel = "Study Week"
#' )
#'
#' # unequal number of subjects across treatments
#' comp_outcome2 <- comp_outcome[
#'   (comp_outcome$trtn == 1 & comp_outcome$usubjid %in% c(1:40)) |
#'     (comp_outcome$trtn == 2 & comp_outcome$usubjid %in% c(101:160)) |
#'     (comp_outcome$trtn == 3 & comp_outcome$usubjid %in% c(201:250)),
#' ]
#' stacked_barchart(
#'   data = comp_outcome2,
#'   chartcolors = colfun()$fig12_colors,
#'   xlabel = "Study Week"
#' )
#'
#' # unequal number of observations across visits
#' comp_outcome3 <- comp_outcome[!(comp_outcome$trtn == 1 &
#'   comp_outcome$usubjid %in% c(1:40) & comp_outcome$visit == 5), ]
#' stacked_barchart(
#'   data = comp_outcome3,
#'   chartcolors = colfun()$fig12_colors,
#'   xlabel = "Study Week"
#' )
stacked_barchart <- function(data, chartcolors, xlabel = "Visit") {
  all_columns <- c(
    "usubjid", "visit", "trt", "brcat"
  )
  nonexistent_columns <- setdiff(all_columns, colnames(data))

  if (length(nonexistent_columns) > 0) {
    error_message <- paste0("You are missing a required variable in your dataframe: ", nonexistent_columns)
    stop(error_message)
  }

  df_n1 <- data %>%
    group_by(trt, visit) %>%
    summarise(n1 = n())

  if (nrow(unique(df_n1[c("trt", "n1")])) > nrow(unique(df_n1["trt"]))) {
    warning(paste(
      "You have unequal number of observations across visits, please check missing data."
    ))
  }

  df_n2 <- data %>%
    group_by(trt, visit, brcat) %>%
    summarise(n2 = n())

  df_stacked <- merge(df_n1, df_n2, by = c("trt", "visit"))
  df_stacked$percentage <- df_stacked$n2 / df_stacked$n1 * 100

  fig <- ggplot(df_stacked, aes(x = visit, y = percentage, fill = brcat)) +
    facet_wrap(~trt, scales = "free_x") +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = chartcolors) +
    geom_text(aes(label = round(percentage, 0)),
      vjust = 1.3, position = "Stack",
      size = control_fonts()$p * 0.35
    ) +
    scale_y_continuous(expand = c(0.015, 0)) +
    xlab(xlabel) +
    ylab("Percentage") +
    guides(fill = guide_legend(title = "Outcome", nrow = 2, byrow = TRUE)) +
    labs(color = NULL) +
    br_charts_theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks.y = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )

  fig
}
```

::: panel-tabset
## Standard Visualization

```{r, echo=FALSE, warning=FALSE, message=FALSE}
stacked_barchart(
  data = comp_outcome,
  chartcolors = colfun()$fig12_colors,
  xlabel = "Study Week"
)
```

## Sample Code

```{r, eval=FALSE}
stacked_barchart(
  data = comp_outcome,
  chartcolors = colfun()$fig12_colors,
  xlabel = "Study Week"
)
```

## Description

**How to read**

-   This visualization is comprised of a separate chart for each dose level placebo, low dose and high dose.

-   The Y-axis displays the percentage of subjects 0 to 100%.

-   The X-axis displays the study visits by study week.

-   Each stacked bar in the display area is comprised of five bars representing the five possible outcomes defined in the legend above the graph.

-   The five ordinal outcomes range from best scenario (e.g. dark blue = benefit without AE) to worst scenario (e.g., dark red = withdrew).

-   The outcome category is determined for each subject at each study week. Once subjects withdraw, they continued to be counted in the “Withdrew” category at subsequent visits.

-   The visualization compares the distribution of the five ordinal outcomes within a visit, across visits, and across doses.

**Key Conclusions:**

-   In this hypothetical example, the placebo control group has more subjects with “No Benefit No AE.” The proportion of subjects with benefits is the highest in the high-dose group.

-   The most favorable outcome of “Benefit-No AE” is not only sustainable but the percentage of subjects is increasing.

-   It is expected that the percentage of subjects withdrawing increase over time, due to placebo subjects receiving no benefit and other subjects experiencing adverse events.

:::
